Empirical Bayes analysis of \RNAseq{} data for detection of gene expression heterosis

### 2015-08-31 response.tex

\comment{Page 3. Bottom.  Shouldn’t you at least provide a justification when one would use a normal prior and when one would or should use a Laplace prior?}

\response{We have assessed the use of normal distributions see Section 3.3 and also provided rationale for why Laplace distributions would be expected to perform better in this scenario at the end of Section 2.1.}

\comment{Page 4, bottom half.  The strategy described here, also somewhat attested to by a reviewer, is a little disconcerting with respect to the “double estimation” of gene-specific parameters…first to estimate they hyperparameters and then again conditioned on the estimates of these hyperparameters.  Perhaps you should characterize that these second-stage gene-specific estimates are shrunk relative to those estimates used to estimate the hyperparameters and that they then have lead to predictions with better properties?  Ideally, I would have liked to see these hyperparameters be also formally estimated together with the gene specific parameters within one MCMC analyses…authors should address why they didn’t do this.}

\response{We now compare the estimated gene-specific parameters to the eBayes estimates in Figure 3. We have added a paragraph to the beginning of Section 2.2 to share our experience with a fully MCMC analysis which took too long and had poor mixing.}

@@

\response{The area under the curve attempts to do exactly this, i.e. a single metric to compare the methods. We do not try to over-emphasize the difference between the ShrinkBayes and eBayes methods, which are both based on our model, as we do not believe there is much of a difference.}

Reviewer \#4

\comment{1. The authors propose a method to assess "gene expression heterosis."   While it is mentioned that this type of heterosis is a possible explanation for phenotypic heterosis, I'd like more discussion regarding this point.  The two references seem a bit outdated (for bioinformatics) so I'm wondering if this explanation is still being considered.  If it is, than I'd like to see more discussion about how the analytic results would be used to investigate this explanation.   Wouldn't scientists have specific phenotypic traits in mind and know of genes associated with these traits?  I'm a little confused why the search through the whole genome for this type of heterosis if they're looking to draw inference on the association.}

\response{Even though scientists may have specific candidates. An exploratory analysis to suggest other genes seems useful.} \todo{Updated references.}

### 2015-08-31 ebayes_rnaseq_heterosis.tex
some changes

\begin{abstract}
Heterosis, or hybrid vigor, refers to the enhancements in the phenotype of hybrid progeny relative to their inbred parents. Although the heterosis phenomenon is extensively utilized in agriculture, the molecular basis of heterosis is still largely unknown. In an effort to understand phenotypic heterosis at the molecular level, researchers have begun to measure transcript abundance levels of thousands of genes in parental inbred lines and their hybrid offspring using RNA sequencing (\RNAseq{}) technology.  The resulting data allow researchers to search for evidence of gene expression heterosis as one potential molecular mechanism underlying heterosis of agriculturally important traits.  The null hypotheses of greatest interest in testing for gene expression heterosis are composite null hypotheses that are difficult to test with standard statistical approaches for \RNAseq{} analysis. To address these shortcomings, we develop a hierarchical negative binomial model and draw inferences using a computationally tractable empirical Bayes approach to inference. We demonstrate improvements over alternative methods via a simulation study based on a maize experiment and then analyze that maize experiment with our newly proposed methodology. 
% This article has supplementary material online.
\end{abstract}

@@

Heterosis exists when the expected value of a hybrid phenotype differs from the average of the expected phenotypic values of the hybrid's parents. The heterosis phenomenon was scientifically documented in plants by \cite{darwin1876effects} and has long been used to improve agricultural production. Overdominance is a particularly extreme case of heterosis where the hybrid phenotype lies outside the range of the parental phenotypes. One classic example involves hybrid maize offspring that are taller, faster to mature, and yield considerably more grain than their inbred parents \citep{hallauer1981quantitative, hallauer2010quantitative}.  Despite intensive study and successful use of heterosis in agriculture, the basic molecular genetic mechanisms remain largely unknown \citep{coors1999genetics, lippman2007heterosis}. One potential explanation, known as gene expression heterosis, is enhanced expression of one or more genes in hybrids compared to their inbred parents \citep{swanson2006all, springer2007allelic}.

@@

The remainder of the paper proceeds as follows. Section \ref{s:method} presents the proposed hierarchical model, an empirical Bayes method of estimating the parameters, and the calculation of posterior probabilities of overdominance. Section \ref{s:simulation} presents a simulation study based on a maize experiment and compares our approach to alternative methods. Section \ref{s:maize} analyzes a maize experiment where phenotypic heterosis is well established and identifies genes demonstrating expression heterosis. The data and code necessary to perform this analysis are available in the Supplemental Material. Section \ref{s:discussion} summarizes the work and suggests directions for future research.

@@

Following \cite{ji2014estimation}, we reparameterize the gene-variety mean structure into the gene-specific parental average ($\phi_g$), half-parental difference ($\alpha_g$), and hybrid difference ($\delta_g$). For our heterosis study where $V=3$, we let $v=1,2$ indicate the two parental varieties and $v=3$ indicate the hybrid. The reparameterization is  
\[ \phi_g = \frac{\mu_{g1}+\mu_{g2}}{2}, \quad 
\alpha_g = \frac{\mu_{g1}-\mu_{g2}}{2}, \quad \mbox{and} \quad
\delta_g = \phi_g-\mu_{g3}. \]

We assume a hierarchical model for the gene-specific mean parameters and overdispersion parameters. Initially, we assume the parental averages, half-parental differences, hybrid effect, and overdispersion parameters follow normal distributions, i.e.
\[ 
%\gamma_{vi} \stackrel{ind}{\sim} N(0,\sigma_\gamma^2), \quad 
\phi_g \stackrel{ind}{\sim} N(\eta_\phi, \sigma_\phi^2), 
\alpha_g \stackrel{ind}{\sim} N(\eta_\alpha, \sigma_\alpha^2),
\delta_g \stackrel{ind}{\sim} N(\eta_\delta, \sigma_\delta^2), \mbox{ and }
\psi_g \stackrel{ind}{\sim} N(\eta_\psi,\sigma_\psi^2). 
\]
Empirical plots of estimated values of $\alpha_g$ and $\delta_g$ for our maize data set (described in Section \ref{s:maize}) suggest that the distribution of these parameters are more peaked near zero and have heavier tails than a normal distribution allows. For differential expression between the two parental phenotypes, we would expect many genes to have small effects and only a few genes to have relatively large effects. For these many genes with small effects, we might expect the hybrid to act similar to its parents and therefore also have many genes where the hybrid effect is small and only a few genes where this hybrid effect is large. For these reasons, we also assessed Laplace (or double exponential) distributions for the half-parental difference and hybrid effect and thus implement a Bayesian LASSO \citep{park2008bayesian}, i.e. 
\[ 
\alpha_g \stackrel{ind}{\sim} La(\eta_\alpha, \sigma_\alpha) \qquad \mbox{and} \qquad
\delta_g \stackrel{ind}{\sim} La(\eta_\delta, \sigma_\delta) 
\]
where $\alpha\sim La(\eta,\sigma)$ has a probability density function given by $La(\alpha;\eta,\sigma) = \exp(-|\alpha-\eta|/\sigma)/2\sigma$ with expectation $\eta$ and variance $2\sigma^2$. Whether using normal or Laplace distributions, we assume \emph{a priori} independence amongst the parental averages, half parental differences, hybrid effects, and overdispersion parameters.

@@

Initial attempts to estimate the models above via Markov chain Monte Carlo (MCMC) failed due to high computational costs and poor mixing in the resulting chains. For example, we implemented the model in the statistical software \Stan{} (discussed at the end of this section), ran the MCMC on a simulated data set with 10,000 genes for 2 months on a state-of-the-art linux server, and obtained potential scale reduction factors that suggested we would need to run at least ten times as long to obtain convergence. Although there are certainly improvements that could be made to decrease computational costs and improve mixing, we opted for an empirical Bayes approach. This approach may be a reasonable approximation to a fully Bayesian approach when estimating models with large numbers of genes as the posterior distributions for the hyperparameters may be tightly peaked.

@@

To obtain normalization factors $\hat{\gamma}$, we use the weighted trimmed mean of $M$ values (TMM) method of \cite{robinson2010scaling}. We then use the methods of \cite{robinson2010edgeR} version 3.8.6, to obtain genewise dispersion estimates, $\hat{\psi}_g$, and their generalized linear model methods to obtain estimates for the remaining gene-specific parameters, i.e. $\hat{\phi}_g$, $\hat{\alpha}_g$, and $\hat{\delta}_g$. Using $\hat{\theta}_g = (\hat{\phi}_g,\hat{\alpha}_g,\hat{\delta}_g,\hat{\psi}_g)$, we estimate hyperparameters for the location and scale parameters in the hierarchical model using a central method of moments approach. For example, $\hat{\eta}_\phi = \sum_{g=1}^G \hat{\phi}_g/G$ and $\hat{\sigma}_\phi^2 = \sum_{g=1}^G (\hat{\phi}_g - \hat{\eta}_\phi)^2/(G-1)$ (and similarly for $\hat{\eta}_\psi$ and $\hat{\sigma}_\psi$) and $\hat{\eta}_\alpha = \sum_{g=1}^G \hat{\alpha}_g/G$ and $2\hat{\sigma}_\alpha^2 = \sum_{g=1}^G (\hat{\alpha}_g - \hat{\eta}_\alpha)^2/(G-1)$ (and similarly for $\hat{\eta}_\delta$ and $\hat{\sigma}_\delta$) for the model assuming Laplace distributions.

@@

In the maize context that motivates this work, we are interested in overdominance, either low-parent heterosis (LPH) or high-parent heterosis (HPH), in gene expression. 
For a specific gene $g$, LPH occurs when average expression in the hybrid is less than the average expression of either parent, i.e. $\delta_g < -|\alpha_g|$ or, equivalently, $\mu_{g3} < \min\{\mu_{g1},\mu_{g2}\}$), and HPH occurs when average expression in the hybrid is greater than the average expression of either parent, i.e. $\delta_g > |\alpha_g|$ or, equivalently, $\mu_{g3} > \max\{\mu_{g1},\mu_{g2}\}$). 
% and high-parent  heterosis (HPH) of gene expression occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$. For simplicity, we define \emph{heterosis} of gene expression to be the event of LPH or HPH. The most important hypotheses are
% \begin{align}
% H_{g0}:&\mu_{g1}\le \mu_{g3}\le \mu_{g2} \mbox{ or } \mu_{g1}\ge \mu_{g3}\ge \mu_{g2} \mbox{\ \ vs.\ \ } \nonumber \\
% H_{g1}:&\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3},
% \label{e:hypotheses}
% \end{align}
% where $H_{g0}$ indicates no heterosis for gene $g$ and $H_{g1}$ indicates  heterosis for gene $g$. 
% We evaluate these hypotheses based on empirical Bayes estimates of their posterior probabilities, e.g., 
% \begin{align}
% P\left(H_{g1}|y, \hat{\pi}_{MAP}\right) &= P\left(\left.\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}\right| y, \hat{\pi}_{MAP}\right) \nonumber \\
% &\approx \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\mu_{g3}^{(m)}< \min\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) \mbox{ or } \max\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) < \mu_{g3}^{(m)}\right), \label{e:probs}
% \end{align}
% where $\mu_g^{(m)} = \left(\mu_{g1}^{(m)},\mu_{g2}^{(m)},\mu_{g3}^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior, and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise. 
We evaluate these probabilities based on empirical Bayes estimates of their posterior probabilities, e.g., 
\begin{equation}
%P(LPH_g|y) \approx 
P(\mbox{LPH}_g|y,\hat{\pi},\hat{\gamma}) = 
%P\left(\left. \mu_{g3} < \min\{\mu_{g1},\mu_{g2}\}\right|y,\hat{\pi},\hat{\gamma}\right) = 
P\left(\left.\delta_g < -|\alpha_g|\right|y, \hat{\pi},\hat{\gamma}\right) = \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\delta_g^{(m)} < -\left|\alpha_g^{(m)}\right|\right), 
\label{e:probs}
\end{equation}
where $\left(\delta_g^{(m)},\alpha_g^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior, and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise. HPH probability is defined similarly with the inequality reversed and without the negative sign. We construct a ranked list of genes according to the maximum of the gene's LPH and HPH heterosis probabilities. Geneticists can use this list to prioritize future experiments to understand the molecular genetic mechanisms for heterosis.

@@

In addition to the approach above, we utilized \ShrinkBayes{} to estimate heterosis probabilities with two modifications described here. \ShrinkBayes{} utilizes integrated nested Laplace approximation (INLA) \citep{rue2009approximate} in combination with empirical Bayes ideas \citep{van2014shrinkbayes}. One limitation with inferential methods based on INLA is that all distributions, except for the data distribution, must have tails as light or lighter than the normal density. Thus, we cannot implement the Laplace priors for the half-parental difference ($\alpha_g$) and the hybrid difference ($\delta_g$) and instead use normal priors in this situation. An additional limitation is that INLA provides approximations to marginal posteriors for parameters or linear combinations of parameters, but not an approximation to the full posterior. Since we are interested in non-linear quantities such as $P(\delta_g > |\alpha_g||y)$, we cannot compute these directly using \ShrinkBayes{}. Instead, for \ShrinkBayes{}, we calculate heterosis probabilities conditional on posterior means for $\hat{\delta}_g$ and $\hat{\alpha}_g$, e.g. 
\[ 
% P_{\ShrinkBayes{}}(LPH_g|y) \equiv \left\{ \begin{array}{ll}
% P(\delta_g>-\alpha_g|y) & \mbox{if } \hat{\delta}_g > -\hat{\alpha}_g \ge 0 \\
% P(\delta_g>\phantom{-}\alpha_g|y) & \mbox{if } \hat{\delta}_g > \phantom{-}\hat{\alpha}_g \ge 0 \\
% 0 & \mbox{otherwise}
% \end{array} \right.
P_{\ShrinkBayes{}}(\mbox{LPH}_g|y) \equiv P(\delta_g<-\alpha_g|y)\mathrm{I}\left(\hat{\delta}_g < -\hat{\alpha}_g \le 0\right) + P(\delta_g<\alpha_g|y) \mathrm{I}\left(\hat{\delta}_g < \hat{\alpha}_g \le 0\right).
%P_{\ShrinkBayes{}}(HPH|y) \equiv \left\{ \begin{array}{ll}
%P(\delta_g<-\alpha_g|y) & \mbox{if } \hat{\delta}_g > -\hat{\alpha}_g \ge 0 \\
%(\delta_g<\phantom{-}\alpha_g|y) & \mbox{if } \hat{\delta}_g > \phantom{-}\hat{\alpha}_g \ge 0 \\
%0 & \mbox{otherwise}
%\end{array} \right.
\]
HPH probability is defined similarly with all inequalities reversed. As before, we construct a ranked list of genes according to the maximum of the gene's LPH and HPH heterosis probabilities.
%the heterosis probability for gene $g$ is zero if $|\hat{\delta}_g| \le |\hat{\alpha}_g|$, $P(\delta_g>\alpha_g|y)$ if $\hat{\delta}_g > -\hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g > 0 |y)$ if $\hat{\delta}_g > \hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g < 0 |y)$ if $\hat{\delta}_g < \hat{\alpha}_g \le 0$, and $P(\delta_g + \alpha_g < 0 |y)$ if $\hat{\delta}_g < -\hat{\alpha}_g \le 0$.

@@

To assess the efficacy of our method to identify genes demonstrating heterosis, we used a maize data set with parental varieties B73 and Mo17 and the hybrid variety (B73 $\times$ Mo17) \citep{paschold2012complementation} to determine realistic parameter values for a simulation study. Section \ref{s:maize} describes the maize dataset in detail. We compared our method to approaches using the R packages \edgeR{}, \baySeq{} \citep{hardcastle2010bayseq, hardcastle2012baySeq}, and \ShrinkBayes{} \citep{van2014shrinkbayes}.

>>>>

\begin{abstract}
Heterosis, or hybrid vigor, refers to the enhancements in the phenotype of hybrid progeny relative to their inbred parents. Although the heterosis phenomenon is extensively utilized in agriculture, the molecular basis of heterosis is still largely unknown. In an effort to understand phenotypic heterosis at the molecular level, researchers are measuring transcript abundance levels of thousands of genes in parental inbred lines and their hybrid offspring using RNA sequencing (\RNAseq{}) technology.  The resulting data allow researchers to search for evidence of gene expression heterosis as one potential molecular mechanism underlying heterosis of agriculturally important traits.  The null hypotheses of greatest interest in testing for gene expression heterosis are composite null hypotheses that are difficult to test with standard statistical approaches for \RNAseq{} analysis. To address these shortcomings, we develop a hierarchical negative binomial model and draw inferences using a computationally tractable empirical Bayes approach to inference. We demonstrate improvements over alternative methods via a simulation study based on a maize experiment and then analyze that maize experiment with our newly proposed methodology. 
% This article has supplementary material online.
\end{abstract}

@@

Heterosis exists when the expected value of a hybrid phenotype differs from the average of the expected phenotypic values of the hybrid's parents. The heterosis phenomenon was scientifically documented in plants by \cite{darwin1876effects} and has long been used to improve agricultural production. Overdominance is a particularly extreme case of heterosis where the expected value of the hybrid phenotype lies outside the range of the expected values of the parental phenotypes. One classic example involves hybrid maize offspring that are taller, faster to mature, and yield considerably more grain than their inbred parents \citep{hallauer1981quantitative, hallauer2010quantitative}.  Despite intensive study and successful use of heterosis in agriculture, the basic molecular genetic mechanisms remain largely unknown \citep{coors1999genetics, lippman2007heterosis}. One potential explanation, known as gene expression heterosis, is enhanced expression of one or more genes in hybrids compared to their inbred parents \citep{swanson2006all, springer2007allelic}.

@@

The remainder of the paper proceeds as follows. Section \ref{s:method} presents the proposed hierarchical model, an empirical Bayes method of estimating the parameters, and the calculation of posterior probabilities of overdominance. Section \ref{s:simulation} presents a simulation study based on a maize experiment and compares our approach to alternative methods. Section \ref{s:maize} analyzes a maize experiment where phenotypic heterosis is well established and identifies genes demonstrating expression overdominance. The data and code necessary to perform this analysis are available in the Supplemental Material. Section \ref{s:discussion} summarizes the work and suggests directions for future research.

@@

Following \cite{ji2014estimation}, we reparameterize the gene-variety mean structure into the gene-specific parental average ($\phi_g$), half-parental difference ($\alpha_g$), and hybrid effect ($\delta_g$). For our heterosis study where $V=3$, we let $v=1,2$ indicate the two parental varieties and $v=3$ indicate the hybrid. The reparameterization is  
\[ \phi_g = \frac{\mu_{g1}+\mu_{g2}}{2}, \quad 
\alpha_g = \frac{\mu_{g1}-\mu_{g2}}{2}, \quad \mbox{and} \quad
\delta_g = \phi_g-\mu_{g3}. \]

We assume a hierarchical model for the gene-specific mean parameters and overdispersion parameters. Initially, we assume the parental averages, half-parental differences, hybrid effect, and overdispersion parameters follow normal distributions, i.e.
\[ 
%\gamma_{vi} \stackrel{ind}{\sim} N(0,\sigma_\gamma^2), \quad 
\phi_g \stackrel{ind}{\sim} N(\eta_\phi, \sigma_\phi^2), \,
\alpha_g \stackrel{ind}{\sim} N(\eta_\alpha, \sigma_\alpha^2), \,
\delta_g \stackrel{ind}{\sim} N(\eta_\delta, \sigma_\delta^2), \, \mbox{ and } \,
\psi_g \stackrel{ind}{\sim} N(\eta_\psi,\sigma_\psi^2). 
\]
Empirical plots of estimated values of $\alpha_g$ and $\delta_g$ for our maize data set (described in Section \ref{s:maize}) suggest that the distribution of these parameters are more peaked near zero and have heavier tails than a normal distribution allows. For differential expression between the two parental phenotypes, we would expect many genes to have small effects and only a few genes to have relatively large effects. For these many genes with small effects, we might expect the hybrid to act similar to its parents and therefore also have many genes where the hybrid effect is small and only a few genes where this hybrid effect is large. For these reasons, we also assessed Laplace (or double exponential) distributions for the half-parental difference and hybrid effect and thus implement a Bayesian LASSO \citep{park2008bayesian,hans2009bayesian}, i.e. 
\[ 
\alpha_g \stackrel{ind}{\sim} La(\eta_\alpha, \sigma_\alpha) \qquad \mbox{and} \qquad
\delta_g \stackrel{ind}{\sim} La(\eta_\delta, \sigma_\delta) 
\]
where $\alpha\sim La(\eta,\sigma)$ has a probability density function given by $La(\alpha;\eta,\sigma) = \exp(-|\alpha-\eta|/\sigma)/2\sigma$ with expectation $\eta$ and variance $2\sigma^2$. Whether using normal or Laplace distributions, we assume \emph{a priori} independence amongst the parental averages, half parental differences, hybrid effects, and overdispersion parameters.

@@

Initial attempts to perform a fully Bayesian analysis via Markov chain Monte Carlo (MCMC) failed due to high computational costs and poor mixing in the resulting chains. For example, we implemented the model in the statistical software \Stan{} (discussed at the end of this section), ran the MCMC on a simulated data set with 10,000 genes for 2 months on a state-of-the-art linux server, and obtained potential scale reduction factors that suggested we would need to run at least ten times as long to obtain convergence. Although there are certainly improvements that could be made to decrease computational costs and improve mixing, we opted for an empirical Bayes approach. This approach may be a reasonable approximation to a fully Bayesian approach when estimating models with large numbers of genes as the posterior distributions for the hyperparameters may be tightly peaked.

@@

To obtain normalization factors $\hat{\gamma}$, we use the weighted trimmed mean of $M$ values (TMM) method of \cite{robinson2010scaling}. We use \edgeR{} to obtain genewise dispersion estimates, $\hat{\psi}_g$, and the generalized linear model methods to obtain estimates for the remaining gene-specific parameters, i.e. $\hat{\phi}_g$, $\hat{\alpha}_g$, and $\hat{\delta}_g$ \citep{robinson2010edgeR}. Using $\hat{\theta}_g = (\hat{\phi}_g,\hat{\alpha}_g,\hat{\delta}_g,\hat{\psi}_g)$, we estimate hyperparameters for the location and scale parameters in the hierarchical model using a central method of moments approach. For example, $\hat{\eta}_\phi = \sum_{g=1}^G \hat{\phi}_g/G$ and $\hat{\sigma}_\phi^2 = \sum_{g=1}^G (\hat{\phi}_g - \hat{\eta}_\phi)^2/(G-1)$ (and similarly for $\hat{\eta}_\psi$ and $\hat{\sigma}_\psi$) and $\hat{\eta}_\alpha = \sum_{g=1}^G \hat{\alpha}_g/G$ and $2\hat{\sigma}_\alpha^2 = \sum_{g=1}^G (\hat{\alpha}_g - \hat{\eta}_\alpha)^2/(G-1)$ (and similarly for $\hat{\eta}_\delta$ and $\hat{\sigma}_\delta$) for the model assuming Laplace distributions.

@@

In the maize context that motivates this work, we are interested in overdominance, either low-parent heterosis (LPH) or high-parent heterosis (HPH), in gene expression. 
For a specific gene $g$, LPH occurs when average expression in the hybrid is less than the average expression of either parent, i.e. $\delta_g < -|\alpha_g|$ or, equivalently, $\mu_{g3} < \min\{\mu_{g1},\mu_{g2}\}$, and HPH occurs when average expression in the hybrid is greater than the average expression of either parent, i.e. $\delta_g > |\alpha_g|$ or, equivalently, $\mu_{g3} > \max\{\mu_{g1},\mu_{g2}\}$. 
% and high-parent  heterosis (HPH) of gene expression occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$. For simplicity, we define \emph{heterosis} of gene expression to be the event of LPH or HPH. The most important hypotheses are
% \begin{align}
% H_{g0}:&\mu_{g1}\le \mu_{g3}\le \mu_{g2} \mbox{ or } \mu_{g1}\ge \mu_{g3}\ge \mu_{g2} \mbox{\ \ vs.\ \ } \nonumber \\
% H_{g1}:&\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3},
% \label{e:hypotheses}
% \end{align}
% where $H_{g0}$ indicates no heterosis for gene $g$ and $H_{g1}$ indicates  heterosis for gene $g$. 
% We evaluate these hypotheses based on empirical Bayes estimates of their posterior probabilities, e.g., 
% \begin{align}
% P\left(H_{g1}|y, \hat{\pi}_{MAP}\right) &= P\left(\left.\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}\right| y, \hat{\pi}_{MAP}\right) \nonumber \\
% &\approx \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\mu_{g3}^{(m)}< \min\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) \mbox{ or } \max\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) < \mu_{g3}^{(m)}\right), \label{e:probs}
% \end{align}
% where $\mu_g^{(m)} = \left(\mu_{g1}^{(m)},\mu_{g2}^{(m)},\mu_{g3}^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior, and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise. 
We evaluate these probabilities based on empirical Bayes estimates of their posterior probabilities, e.g., 
\begin{equation}
%P(LPH_g|y) \approx 
P(\mbox{LPH}_g|y,\hat{\pi},\hat{\gamma}) = 
%P\left(\left. \mu_{g3} < \min\{\mu_{g1},\mu_{g2}\}\right|y,\hat{\pi},\hat{\gamma}\right) = 
P\left(\left.\delta_g < -|\alpha_g|\right|y, \hat{\pi},\hat{\gamma}\right) = \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\delta_g^{(m)} < -\left|\alpha_g^{(m)}\right|\right), 
\label{e:probs}
\end{equation}
where $\left(\delta_g^{(m)},\alpha_g^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior, and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise. HPH probability is defined similarly with the inequality reversed and without the negative sign. We construct a ranked list of genes according to the maximum of the gene's LPH and HPH heterosis probabilities. Geneticists can use this list to prioritize future experiments to understand the molecular genetic mechanisms for heterosis.

@@

In addition to the approach above, we utilized \ShrinkBayes{} to estimate heterosis probabilities with two modifications described here. \ShrinkBayes{} utilizes integrated nested Laplace approximation (INLA) \citep{rue2009approximate} in combination with empirical Bayes ideas \citep{van2014shrinkbayes}. One limitation with inferential methods based on INLA is that all distributions, except for the data distribution, must have tails as light or lighter than the normal density. Thus, we cannot implement the Laplace priors for the half-parental difference ($\alpha_g$) and the hybrid effect ($\delta_g$) and instead use normal priors in this situation. An additional limitation is that INLA provides approximations to marginal posteriors for parameters or linear combinations of parameters, but not an approximation to the full posterior. Since we are interested in non-linear quantities such as $P(\delta_g > |\alpha_g||y)$, we cannot compute these directly using \ShrinkBayes{}. Instead, for \ShrinkBayes{}, we calculate heterosis probabilities conditional on posterior means for the half-parental difference and hybrid effect, i.e. $\hat{\delta}_g$ and $\hat{\alpha}_g$. For example,  
\[ 
% P_{\ShrinkBayes{}}(LPH_g|y) \equiv \left\{ \begin{array}{ll}
% P(\delta_g>-\alpha_g|y) & \mbox{if } \hat{\delta}_g > -\hat{\alpha}_g \ge 0 \\
% P(\delta_g>\phantom{-}\alpha_g|y) & \mbox{if } \hat{\delta}_g > \phantom{-}\hat{\alpha}_g \ge 0 \\
% 0 & \mbox{otherwise}
% \end{array} \right.
P_{\ShrinkBayes{}}(\mbox{LPH}_g|y) \equiv P(\delta_g<-\alpha_g|y)\mathrm{I}\left(\hat{\delta}_g < -\hat{\alpha}_g \le 0\right) + P(\delta_g<\alpha_g|y) \mathrm{I}\left(\hat{\delta}_g < \hat{\alpha}_g \le 0\right).
%P_{\ShrinkBayes{}}(HPH|y) \equiv \left\{ \begin{array}{ll}
%P(\delta_g<-\alpha_g|y) & \mbox{if } \hat{\delta}_g > -\hat{\alpha}_g \ge 0 \\
%(\delta_g<\phantom{-}\alpha_g|y) & \mbox{if } \hat{\delta}_g > \phantom{-}\hat{\alpha}_g \ge 0 \\
%0 & \mbox{otherwise}
%\end{array} \right.
\]
HPH probability is defined similarly with all inequalities reversed. As before, we construct a ranked list of genes according to the maximum of the gene's LPH and HPH heterosis probabilities.
%the heterosis probability for gene $g$ is zero if $|\hat{\delta}_g| \le |\hat{\alpha}_g|$, $P(\delta_g>\alpha_g|y)$ if $\hat{\delta}_g > -\hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g > 0 |y)$ if $\hat{\delta}_g > \hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g < 0 |y)$ if $\hat{\delta}_g < \hat{\alpha}_g \le 0$, and $P(\delta_g + \alpha_g < 0 |y)$ if $\hat{\delta}_g < -\hat{\alpha}_g \le 0$.

@@

To assess the efficacy of our method to identify genes demonstrating heterosis, we used a maize data set with parental varieties B73 and Mo17 and the hybrid variety (B73 $\times$ Mo17) \citep{paschold2012complementation} to determine realistic parameter values for a simulation study. Section \ref{s:maize} describes the maize dataset in detail. We compared our method to approaches using the R packages \edgeR{} , \baySeq{} \citep{hardcastle2010bayseq, hardcastle2012baySeq}, and \ShrinkBayes{} \citep{van2014shrinkbayes}.
